Proposed Solution

### 2017-01-22 CAV17/secs/trans.tex
mostly done with prob types
TODO revisit this to fix so called ambiguity in rules

The translator takes as input a sample dataset of configuration files and transforms it into a typed and well-structured intermediate representation.
The translator can be seen a parser used to generate an intermediate representation for learning rules (see $\S$\ref{sec-learn}).
Translating or parsing is system dependent since each configuration language (MySQL, Apache, PHP) uses a different grammar.
\app allows users to provide extra help to the translator for their specific system configurations.

@@

\begin{figure}[!htb]
    \centering
    \begin{minipage}{.25\textwidth}
	\begin{lstlisting} [label={lst:1.cnf},language=C,caption={1.cnf}]
[server]
foo = ON
[client]
bar = 1
	\end{lstlisting}
    \end{minipage}%
    \hspace{1cm}
    \begin{minipage}{0.25\textwidth}
	\begin{lstlisting} [label={lst:2.cnf},language=C,caption={2.cnf}]
[server]
foo = ON
[client]
bar = ON
	\end{lstlisting}
    \end{minipage}
    \hspace{1cm}
    \begin{minipage}{0.25\textwidth}
	\begin{lstlisting} [label={lst:2.cnf},language=C,caption={3.cnf}]
[server]
foo = OFF
[client]
bar = OFF
	\end{lstlisting}
    \end{minipage}

    \caption{A learning set of configuration files}
\end{figure}

An additional challenge is that it is not always possible to fully determine the type of key based on a single example value. 
For this reason, we introduce {\emph {probabilistic types}}.
Taking the configuration \ref{lst:1.cnf}, we can assume {\tt foo} is a boolean type by the grammar of MySQL,
  but the keyword {\tt bar} could either be a integer type or a Boolean type.
Consider that we would like to learn a rule that the boolean keywords should have the same values ($eq(\texttt{foo},\texttt{bar})$).
We then have two potential situations; in the first case where {\tt bar:int}, we do not learn the above rule, but in the second case where {\tt bar:bool}, the rule is valid.
To resolve this ambiguity, instead of immediately assigning one type to a value, the translator assigns a distribution of types to a value based on examples from the learning set.
% an idea closely related to existentially quantified types~\cite{Launchbury93lazyfunctional}.

Formally, we define probabilistic types as follows: let $\mathcal{T}$ be a set of basic types.
In \app, the set $\mathcal{T}$ contains strings, file paths, integers, sizes, and booleans. 
A probabilistic type, $\tilde{\tau} \in \tilde{\mathcal{T}}$, is a set of pairs $\{(\tau_1, c_1),\ldots,(\tau_n, c_n)\}$, such that $\tau_i \in \mathcal{T}$, $c_i \in \mathbb{Z}$. 
A keyword may have a probabilistic type, expressed $k:\tilde{\tau}$.
The count for $\tau_i$ is incremented each time a new example value that might be type $\tau_i$ is encountered for a particular key.

\begin{mathpar}
\hspace{-0.4cm} %no idea why, but this is better than \centering here
\inferrule* [Right=bool\_int]
{(k,v) \in C \\ v \in \{0,1\}}
{k:[int\pluseq 1,bool\pluseq 1]}
\and
\inferrule* [Right=eq\_rule]
{k1 :: \tau \\ k2 :: \tau \\ \tau \in \mathcal{T}}
{eq(k1,k2) :: Rule}
\and
\inferrule* [Right=bool\_resolve]
{k:\tilde{\tau} \\ \tilde{\tau} \vdash bool \geq 2 \land \ast \leq 1}
{k :: bool}
\and
\\
\inferrule* [Right=bool]
{(k,v) \in C \\ v \in \{ON,OFF\}}
{k:[bool\pluseq 1,string\pluseq 1]}
\end{mathpar}

In the {\scriptsize BOOL\_RESOLVE} rule, a keyword with a probabilistic type $k : \tilde{\tau} \in \tilde{\mathcal{T}}$ can be resolved to a basic type (expressed with the double colon $ k :: \tau \in \mathcal{T}$) with enough evidence.
The definition of enough evidence must be empirically determined by the user depending on the quality of the learning set.
In this case, we allow a probabilistic type to resolve to a boolean if we have two counts for boolean, and no more than one count for any other type.

A user may implement overlapping inference rules, for example in {\scriptsize INT\_RESOLVE}.
This not a subtyping relation, as the predicates over the counts may not be strictly overlapping.
We simply allow the double colon (::) to type a single keyword in multiple ways.
Because the only typing judgments on a basic type are terminal, {\scriptsize EQ\_RULE}, there is no problem with ambiguity as we would have in a traditional lambda calculus that included function application.

\begin{mathpar}
%\centering 
\inferrule* [Right=int\_resolve]
{k:\tilde{\tau} \\ \tilde{\tau} \vdash int \geq 1}
{k :: int}
\end{mathpar}

Note that we do not have an application rule over probabilistic types - types must be resolved to learn a rule over a keyword.
In the case that there is not enough evidence to resolve a type, no type-dependent rules may be learned over that keyword.

Taking the example type inference rules above, we can run an example inference.
For example, if the training set contained [(``bar",1),(``bar",ON)], the resulting type would be bar::[bool=2,int=1] (considering that the literal 1 may act as either an int or a boolean).

>>>>

The translator takes as input a training set of configuration files and transforms it into a typed and well-structured intermediate representation.
The translator can be seen a parser used to generate an intermediate representation for learning rules (see $\S$\ref{sec-learn}).
Translating or parsing is system dependent since each configuration language (MySQL, Apache, PHP) uses a different grammar.
\app allows users to provide extra help to the translator for their specific system configurations.

@@

\begin{figure}[!htb]
    \centering
    \begin{minipage}{.25\textwidth}
	\begin{lstlisting} [label={lst:1.cnf},language=C,caption={1.cnf}]
[server]
foo = ON
[client]
bar = 1
	\end{lstlisting}
    \end{minipage}%
    \hspace{1cm}
    \begin{minipage}{0.25\textwidth}
	\begin{lstlisting} [label={lst:2.cnf},language=C,caption={2.cnf}]
[server]
foo = ON
[client]
bar = ON
	\end{lstlisting}
    \end{minipage}
    \hspace{1cm}
    \begin{minipage}{0.25\textwidth}
	\begin{lstlisting} [label={lst:2.cnf},language=C,caption={3.cnf}]
[server]
foo = OFF
[client]
bar = OFF
	\end{lstlisting}
    \end{minipage}
    \caption{A learning set of configuration files}
    \label{fig:tset}
\end{figure}

An additional challenge is that it is not always possible to fully determine the type of key based on a single example value. 
For this reason, we introduce \textit{probabilistic types}, as contrasted with \textit{basic types}.
In \app, the set of basic types contains strings, file paths, integers, sizes, and booleans. 
Taking the configuration \ref{lst:1.cnf}, we can assume {\tt foo} is a boolean type by the grammar of MySQL,
  but the keyword {\tt bar} could either be a integer type or a Boolean type.
Consider that we would like to learn a rule that the boolean keywords should have the same values ($eq(\texttt{foo},\texttt{bar})$).
We then have two potential situations; in the first case where {\tt bar}$::int$, we do not learn the above rule, but in the second case where {\tt bar}$::bool$, the rule is valid.
To resolve this ambiguity, instead of immediately assigning one type to a value, the translator assigns a distribution of types to a value based on examples from the learning set.
% an idea closely related to existentially quantified types~\cite{Launchbury93lazyfunctional}.

Formally, a probabilistic types is a set of counts over a set $\mathcal{T}$ of basic types.
A probabilistic type, $\tilde{\tau} \in \tilde{\mathcal{T}}$, is a set of pairs $\{(\tau_1, c_1),\ldots,(\tau_n, c_n)\}$, such that $\tau_i \in \mathcal{T}$, $c_i \in \mathbb{Z}$. 
A keyword has a probabilistic type, expressed $k:\tilde{\tau}$, which without any evidence remains as {\scriptsize EMPTY}.
The count for $(\tau_i,c_i) \in \tilde{\tau}$ is incremented each time a new example value that matches to type $\tau_i$ is encountered for a particular key, as shown in the {\scriptsize BOOL\_INT} and {\scriptsize BOOL} rules.

\begin{mathpar}
\hspace{-0.4cm} %no idea why, but this is better than \centering here
\inferrule* [Right=Empty]
{\\}
{k:[\ast = 0]}
\and
\inferrule* [Right=bool\_int]
{(k,v) \in C \\ v \in \{0,1\}}
{k:[int\pluseq 1,bool\pluseq 1]}
\\
\and
\inferrule* [Right=bool\_resolve]
{k:\tilde{\tau} \\ \tilde{\tau} \models bool \geq 2 \land \ast \leq 1}
{k :: bool}
\\
\and
\inferrule* [Right=bool]
{(k,v) \in C \\ v \in \{ON,OFF\}}
{k:[bool\pluseq 1]}
\and
\inferrule* [Right=eq\_rule]
{k1 :: \tau \\ k2 :: \tau \\ \tau \in \mathcal{T}}
{eq(k1,k2) :: Rule}
\end{mathpar}

In the {\scriptsize BOOL\_RESOLVE} rule, a keyword with a probabilistic type $k : \tilde{\tau} \in \tilde{\mathcal{T}}$ can be resolved to a basic type with enough evidence.
The definition of enough evidence must be empirically determined by the user depending on the quality of the learning set.
In this case, we allow a probabilistic type to resolve to a boolean if we have two counts for boolean, and no more than one count for any other basic type ($\ast$).
We use the notation, $\tilde{\tau} \models p$, to denote a probabilistic type whose counts hold on the predicate $p$.
Taking the type inference rules above, we can run an example inference on the sample training set in Figure~\ref{fig:tset}.
We would derive the probabilistic type \texttt{bar}$:[bool=3,int=1]$, which then resolves to \texttt{bar}$::bool$.

A user may implement overlapping inference rules, for example in {\scriptsize INT\_RESOLVE}.
This not a subtyping relation, as the predicates over the counts may not be strictly overlapping.
%TODO this make no sense - address ambiguity more clearly
We simply allow the double colon (::) to type a single keyword in multiple ways.
Because the only typing judgments on a basic type are terminal, {\scriptsize EQ\_RULE}, there is no problem with ambiguity as we would have in a traditional lambda calculus that included function application.

\begin{mathpar}
\hspace{-0.8cm} %no idea why, but this is better than \centering here
\inferrule* [Right=int\_resolve]
{k:\tilde{\tau} \\ \tilde{\tau} \models int \geq 1}
{k :: int}
\and
\inferrule* [Right=order\_rule]
{k1,k2 \in C \\ k1 \neq k2}
{ord(k1,k2) :: Rule}
\end{mathpar}

In the case that there is not enough evidence to resolve a probabilistic type, no type-dependent rules may be learned over that keyword.
We are still able to learn rules such as {\scriptsize ORDER}, which do not depend on resolved typed.
